Pushdown Control-Flow Analysis via Refunctionalization (Functional Pearl)

### 2018-03-11 paper.tex
update

Traditionally, continuation-passing style (CPS) is a popular intermediate representation
for analyzing functional programs because of it exposes control transfer explicitly
and simplifies the analysis \cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007}.
Here, we choose to use administrative normal form (ANF) which is a direct-style intermediate representation 
as target for clarity but without losing simplicity and generality.
Although we just show the core calculus language, but it can be easily extended
to support recursive bindings (such as @letrec@), conditionals, primitive types and 
operations on primitive types. These cases would be trival or simple to implement, 
so we omit them in the paper.
The transformations we will show in the rest of this paper
also work on abstract machines for plain direct-style $\lambda$-calculus languages.

@@

\subsubsection{Allocating Addresses}
Up to now, we did not describe much about the adresses in stores and the 
time-stamp @Time@.
But they are key ingredients to achieve analyses with different sensitivities,
as well as a finite state space analysis.
To effectively approximate the runtime behavior, a finite 
program contour @time@ that encodes the program execution history is introduced.
Here, we use a list of execution contexts (expressions) to reprent that, and
as we will see in section \ref{kcfainst}, by using different @tick@
function on time we may obtain a family of analyses.

@@

As we mentioned previously, the space of addresses is finite in AAM.
Binding addresses are parameterized by variable names and the time that creates this binding,
which are both finite.
Continuation addresses @KAddr@ has two variants: 
1) @Halt@ which corresponds to the empty stack, and
2) @ContAddr@ consists of the entry expressions of callee, which are also also
finite.
One may ask, why not also keep track of the time in continuation addresses?
(TODO: why target expression, allocation polyvariance)

@@

\section{Fusing}

@@

\section{Disentangling}

@@

\section{Refunctionalization}

@@

%probably Store Widening \& Abstract Garbage Collection
%monadic
\section{Monadification: to Abstracting Definitional Interpreter}

For one direction, we can transform the refunctionalized AAM to abstracting definitional interpreter
by adding monads.

\section{Un-CPS: to Yet Another Abstracting Definitional Interpreter}

We can transform the refunctionalized AAM to direct style, but explicitly use 
side effects such as assignment and mutation to handle collecting semantics and non-determinism.

TODO: use delimited continuation? 
Another form of abstracting definitional interpreter.

@@

\citeauthor{gilray2016pushdown} proposed a continuation address allocator for 
small-step abstracting abstract machines that is both simple to implement and
computational inexpensive. 
Based on the AAM we presented in Section \ref{aam}, the only change in code is
not only keeping track of the entry-point expression of callee, but also holding the target environment
when allocating a continuation address.
The target environment is the binding environment extended from closure's environment
(namely @val new_env = env_c + (v -> baddr)@). All the other part of the code remain the same.
The following code shows the implementation:

@@

TODO: connection to this paper.
TODO: delimited control vs monad. correspondence since the very beginning time.
TODO: although they establish pushdown control-flow analysis in different ways, 
is there a correspondence between different methods?

>>>>

Traditionally, continuation-passing style (CPS) is a popular intermediate representation
for analyzing functional programs because of it exposes control transfer explicitly
and simplifies the analysis \cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007}.
Here, we choose to use $\lambda$-calculus in administrative normal form (ANF) 
which is a direct-style intermediate representation as target for clarity 
but without losing simplicity and generality.
Although we just show the core calculus language, but it can be easily extended
to support recursive bindings (such as @letrec@), conditionals, primitive types and 
operations on primitive types. These cases would be trival or simple to implement, 
so we omit them in the paper.
The transformations we will show in the rest of this paper
also work on abstract machines for plain direct-style $\lambda$-calculus languages.

@@

\subsubsection{Allocating Addresses}
Up to now, we did not describe much about allocating adresses in stores and the 
time-stamp @Time@.
But in abstract interpretation, they are key ingredients to achieve 
analyses with different sensitivities, as well as to a finite state space analysis\cite{Gilray:2016:ACP:2951913.2951936}.
To effectively approximate the runtime behavior, a finite 
program contour @time@ that encodes the program execution history is introduced.
Here, we use a list of execution contexts (expressions) to represent that, and
as we will see in section \ref{kcfainst}, by using different @tick@
function on time we may obtain a family of analyses.

@@

As we mentioned previously, the space of addresses has to be finite in AAM.
Binding addresses are parameterized by variable names and the time that creates this binding,
which are both finite.
Continuation addresses @KAddr@ has two variants: 
1) @Halt@ which corresponds to the empty stack, and
2) @ContAddr@ consists of the entry expressions of callee, which are also also
finite.
One may ask, why not also keep track of the time in continuation addresses?
(TODO: why target expression, allocation polyvariance)

@@

\section{Fusing} \label{fusing}

@@

\section{Disentangling} \label{disen}

@@

\section{Refunctionalization} \label{refunc}

@@

%probably Store Widening \& Abstract Garbage Collection
\section{To Direct-Style}

In previous section, we presented a refunctionalized AAM that in extended continuation-passing style. 
To obtain a definitional abstract interpreter, we need to transform it further to direct-style. 
At this point, we have two choices: 
1) By monadifying the continuations into monads, we obtain the \citeauthor{darais2017abstracting}'s 
abstracting definitional interpreter in monadic style.
2) By representing the extended continuation-passing style with delimited continuations,
we derive a new form of abstract interpreters in direct-style.
This coincidence should not be a surprise since there are extraordinarily rich literatures 
show the correspondence between monads and continuations as of their very beginning era
(TODO: cite Danvy/abs control/esce of fp).
In this section, we present both of them.

\subsection{Monadification} \label{monadify}

In this section, we show how to transform the refunctionalized AAM to monadic-style
abstracting definitional interpreters.

\subsection{Un-CPS} \label{uncps}

On another path, here, we introduce direct-style abstracting definitional interpreter 
with delimited control operators @reset@ and @shift@.

collecting semantics and non-determinism.

TODO: use delimited continuation?

@@

\citeauthor{gilray2016pushdown} proposed a polyvariant address allocator for continuations
under small-step abstracting abstract machines that is both simple to implement and
computational inexpensive. 
Based on the AAM we presented in Section \ref{aam}, the only change in code is
not only keeping track of the entry-point expression of callee, but also holding the target environment
when allocating a continuation address.
The target environment is the binding environment extended from closure's environment
(namely @val new_env = env_c + (v -> baddr)@). All the other part of the code remain the same.
The following code shows the implementation:

@@

composable.
TODO: connection to this paper.
TODO: although they establish pushdown control-flow analysis in different ways, 
is there a correspondence between different methods?
